Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-07-23 exp1/iswc2013.tex
Some minor fixes
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@612 43ef1190-90fd-4dd7-a769-d0121f02b80e

First, note that the concept of parameters as considered above is
known as \emph{factors} in DoE. For each factor, one may make a
measurement of a \emph{response variable} at certain \emph{levels},
e.g. one may measure the response time of a SPARQL endpoint by
measuring it when it contains 1 or 2 million triples. Then, the
response time is the response variable, the factor is the number of
triples and there are two levels, 1 or 2 million triples. We have chosen in
this paper to only deal with the relatively straightforward aspects of
the DoE formalism, and therefore constrain the experiment strictly to
2 levels.

@@

The next step is to understand which factors are important or
significant.  One approach is to plot the sorted effects against the
normal distribution. To understand why, note that if there is nothing
of interest, the measurements have a certain normal distribution
purely due to noise. When plotted against a normal distribution, the
plot would be a straight line. If there are any deviations from the
straight line, it implies that they may be significant. Such plots can
be found in
figures~\ref{fig:fullnormal},~\ref{fig:frac32normal}~and~\ref{fig:frac64normal}. The
range in the $y$-axis is dictated by the number of runs.

@@

In the full factorial experiment, all combinations of the 8 factors
are executed, in total 256 runs. This is called a $2^8$ factorial
experiment, and we can see that the number of runs grows exponentially
with the number of factors, thus making large experiments
prohibitively expensive. We shall nevertheless find the following
small example instructive. Note that each combination is executed only
once (i.e. it is \emph{unreplicated}), however, so even a full
factorial experiment compares well to a typical benchmark in which
each microbenchmark must be executed several times to obtain enough
data to compute a mean with reasonable certainty.

The above experimental setup is executed across a design matrix
generated with the DoE.base package, which returns a data frame
consisting of the factors with a column containing the corresponding
curl measurements. Then, we
generate a normal plot, see Figure~\ref{fig:fullnormal}. In this plot,
if the null hypothesis $H_0$ holds, all estimated effects will fall on
a straight line, as their means would be zero. Any significant
departure from a straight line can be interpreted as a significant
effect (see Section~4.8 of \cite{wu2009experiments} for a detailed
explanation). In our case, it is most important to note that any
effect that is negative means the effect enhances the performance of
the SPARQL endpoint, the runtime decreases. If ``Implement'' and its
interactions are negative conditional on that the factors involved
with its interactions are on their high level, then it supports the
alternative hypothesis, i.e. we have successfully improved the
performance.

>>>>

First, note that the concept of parameters as considered above is
known as \emph{factors} in DoE. For each factor, one may make a
measurement of a \emph{response variable} at certain
\emph{levels}. For example, one may measure the response time of a
SPARQL endpoint by measuring it when it contains 1 or 2 million
triples. Then, the response time is the response variable, the factor
is the number of triples and there are two levels, 1 or 2 million
triples. We have chosen in this paper to only deal with the relatively
straightforward aspects of the DoE formalism, and therefore constrain
the experiment strictly to 2 levels.

@@

The next step is to understand which factors are important or
significant.  One approach is to plot the sorted effects against the
normal distribution. To understand why, note that if there is nothing
of interest, the measurements have a certain normal distribution
purely due to noise. When plotted against a normal distribution, the
plot would be a straight line. If there are any deviations from the
straight line, it implies that they may be significant. Such plots can
be found in
figures~\ref{fig:fullnormal},~\ref{fig:frac32normal},~\ref{fig:frac64normal}~and~\ref{fig:frac64hugenormal}. The
range in the $y$-axis is dictated by the number of runs.

@@

In the full factorial experiment, all combinations of the 8 factors
are executed, in total 256 runs. This is called a $2^8$ factorial
experiment, and while it quickly becomes infeasible for many factors,
we shall nevertheless find the following small example
instructive. Note that each combination is executed only once (i.e. it
is \emph{unreplicated}), however, so even a full factorial experiment
compares well to a typical benchmark in which each microbenchmark must
be executed several times to obtain enough data to compute a mean with
reasonable certainty.

The above experimental setup is executed across a design matrix
generated with the DoE.base package, which returns a data frame
consisting of the factors with a column containing the corresponding
curl measurements. With that, we do as described in the introduction
and fit a linear model with all factors. Then, we generate a normal
plot, see Figure~\ref{fig:fullnormal}.  Any significant departure from
a straight line can be interpreted as a significant effect (see
Section~4.8 of \cite{wu2009experiments} for a detailed
explanation). In our case, it is most important to note that any
effect that is negative means the effect enhances the performance of
the SPARQL endpoint, the runtime decreases. If ``Implement'' and its
interactions are negative conditional on that the factors involved
with its interactions are on their high level, then it supports the
alternative hypothesis, i.e. we have successfully improved the
performance.
